---
title: "p8105_hw3_mgw2144"
output: github_document
date: "2023-10-11"
---

```{r}
library(tidyverse)
library(ggridges)
library(patchwork)
library(p8105.datasets)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  out.width = "90%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = ("viridis"))

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1 

```{r}
data("instacart")

instacart = 
  instacart %>%
  as_tibble()
```
This dataset contains `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row resprenting a single product from an instacart order. Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. In total, there are `r instacart %>%  select(product_id) %>%  distinct() %>%  count()` products found in `r instacart %>%  select(user_id, order_id) %>%  distinct() %>%  count()` orders from `r instacart %>%  select(user_id) %>%  distinct() %>%  count()` distinct users.

Below is a table summarizing the number of items ordered from aisle. In total, there are 134 aisles, with fresh vegetables and fresh fruits holding the most items ordered by far.

```{r}
instacart %>%  
  count(aisle) %>%  
  arrange(desc(n))
```

Next is a plot that shows the number of items ordered in each aisle. Here, aisles are ordered by ascending number of items.

```{r}
instacart %>%  
  count(aisle) %>% 
  filter(n > 10000) %>% 
  mutate(aisle = fct_reorder(aisle, n)) %>% 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```
Our next table shows the three most popular items in aisles `baking ingredients`, `dog food care`, and `packaged vegetables fruits`, and includes the number of times each item is ordered in your table.

```{r}
instacart %>% 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>% 
  group_by(aisle) %>% 
  count(product_name) %>%  
  mutate(rank = min_rank(desc(n))) %>% 
  filter(rank < 4) %>%  
  arrange(desc(n)) %>% 
  knitr::kable()
```

Finally is a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week. This table has been formatted in an untidy manner for human readers. Pink Lady Apples are generally purchased slightly earlier in the day than Coffee Ice Cream, with the exception of day 5.

```{r}
instacart %>% 
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>% 
  group_by(product_name, order_dow) %>% 
  summarize(mean_hour = mean(order_hour_of_day)) %>% 
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) %>% 
  knitr::kable(digits = 2)
```

## Problem 2 

```{r}
data("brfss_smart2010")

brfss_smart2010_df = 
  brfss_smart2010 %>% 
  janitor::clean_names() %>% 
  rename(
    state = locationabbr,
    state_county = locationdesc
  ) %>% 
  filter(topic == "Overall Health") %>% 
  filter(response %in% c("Excellent", "Very good", "Good", "Fair", "Poor")) %>% 
  mutate(
    response = factor(response, levels = c("Poor", "Fair", "Good", "Very good", "Excellent"), ordered = TRUE)) %>% 
  arrange(response)

```

Data was cleaned to re-format some variable names and clean the column names, focus on the "Overall Health" topic, include only participants whose answers ranged from "Excellent" to "Poor", and so that the responses were organized from "Poor" to "Excellent". 

```{r}
brfss_2002 = 
  brfss_smart2010_df %>% 
  filter(year == "2002") %>% 
  count(state) %>% 
  filter(n >= 7)

brfss_2010 = 
  brfss_smart2010_df %>% 
  filter(year == "2010")%>% 
  count(state) %>% 
  filter(n >= 7)
```

In 2002, 21 states were observed at 7 or more locations including: CT, DE FL, ID, IL, IN, MA, MD, MI, NE, NH, NJ, NV, NY, OH, OK, PA, RI, SC, UT, and WA. 

In 2010, 38 states were observed at 7 or more locations including: AL, AR, AZ, CA, CO, DE FL, ID, IN, KS, LA, MA, MD, ME, MI, MN, MO, MT, NC, NE, NH, NJ, NM, NV, NY, OH, OK, OR, PA, RI, SC, SD, TN, TX, UT, VT, WA, and WY.

```{r}
brfss_exc = 
  brfss_smart2010_df %>% 
  filter(response == "Excellent") %>% 
  select(year, state, response, data_value) %>% 
  group_by(year, state) %>% 
  summarise(mean_dv = mean(data_value, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = mean_dv, color = state)) +
    geom_point() + 
    geom_line() +
    labs(
      title = "Average Data Value Per State 2002-2010",
      x = "Year",
      y = "Average Data Value"
    )

brfss_exc
 
```

The result of this code is a "spaghetti" plot with one line for each state in the data set as is represented by the state abbreviation. Each data point connected by lines represents the average data value in percent for each year by state. The average data value is represented by the y-axis and the range of the graph on the y-axis is 10%-35%. Year is represented on the x-axis with a range of 2002-2010. There appears to be two states with significant dips in average data value with one seeing a drop in 2005 and 2009, and the other seeing a dip in 2007.  

```{r}
brfss_two = 
 brfss_smart2010_df%>% 
  filter(
    state == "NY", year %in% c(2006, 2010)) %>% 
  mutate(
    state_county = sub(" County", "", state_county),
    state_county = sub("NY - ", "", state_county)
  ) %>% 
ggplot(aes(x = response, y = data_value, fill = response))+
  geom_bar(stat = "identity")+
  facet_grid(state_county~year)+
  labs(
    x = "Response",
    y = "Average Data Value (%)",
    title = "Average Data Value For New York Counties in 2006 and 2010 by Response Type"
  )

brfss_two
```

This plot reflects the distribution of average percent data value by response category for the years 2006 and 2010 for NY counties. Overall, the "Poor" response category appears to have the lowest average data value for every county in both years surveyed. The "Good" and "Very good" categories appears to have the greatest average data values for each county generally for both 2006 and 2010. More counties were surveyed in 2010 than in 2006 which shows an increase in survey range between the two survey periods. All distributions appear to be left-skewed with the exception of the Bronx county in 2010 which may be slightly more symmetrical. 

## Problem 3

```{r}
nhanes_accel =
  read_csv("C:/Users/magno/Desktop/p8105_hw3_mgw2144/data/nhanes_accel.csv") %>% 
  janitor::clean_names()

nhanes_covar =
  read_csv("C:/Users/magno/Desktop/p8105_hw3_mgw2144/data/nhanes_covar.csv", skip = 4) %>% 
  janitor::clean_names() %>% 
  drop_na()

nhanes = 
  nhanes_accel %>% 
  inner_join(nhanes_covar, by = "seqn") %>% 
  relocate(seqn, sex, age, bmi, education) %>% 
  filter(age >= 21) %>% 
  mutate(
    sex = case_when(
      sex == 1 ~ "male",
      sex == 2 ~ "female"),
    education = case_when(
      education == 1 ~ "less than high school",
      education == 2 ~ "high school equivalent",
      education == 3 ~ "more than high school")) %>% 
  mutate(
    sex = forcats::fct_relevel(sex, "male", "female"),
    education = forcats::fct_relevel(education, "less than high school", "high school equivalent", "more than high school")
  )
```

```{r}
sex_education = 
  nhanes %>% 
  select(sex, education) %>% 
  group_by(sex, education) %>% 
  summarize(count = n()) 

sex_education

sex_age_education = 
  nhanes %>% 
  select(sex, education, age) %>% 
  ggplot(aes(x = age, fill = sex))+
  geom_histogram(position = "dodge", binwidth = 5) +
  facet_grid(sex~education) +
  labs(title = "Age Distribution by Sex and Education Level", x = "Age", y = "Count") 

sex_age_education
```

The sex_education table shows the number of men and women in each category by education level laid out in a 6x3 dataframe.

This plot illustrates distribution of individuals in each education category by age for individuals over the age of 21 separated by sex. In the male category there appear to be the greatest number of individuals in the "more than high school" category and the fewest number of individuals in the "less than high school" category. In the female category there appear to be the greatest number of individuals in the "more than high school" category and the fewest number of individuals in the "high school equivalent" category. The female "more than high school" distribution appears to have a strong right skew with the majority of females in this distribution being younger that reported "more than high school". In the "high school equivalent" category there is a higher representation of males than females. 

```{r warning=FALSE} 
nhanes_act_total =
  nhanes %>% 
  rowwise() %>% 
  mutate(
    total_activity = sum(c_across(min1:min1440)) %>% 
    as.numeric())

total_act_plot = 
  nhanes_act_total %>% 
  ggplot(aes(x = age, y = total_activity, color = total_activity)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_grid(sex~education) +
  labs(title = "Total Activity and Age by Sex and Education Level", x = "Age", y = "Total Activity in Minutes")

total_act_plot
```

This plot illustrates the total activity for a 24 hours period in minutes by age stratified by reported education level and sex for adults aged 21 and older. Among all sex and education categories it appears that total daily activity minutes decreases with age overall. With the exception of females in the "high school equivalent" category and males in the "less than highschool" category, the relationship between total activity in minutes and age appears to be fairly linear and negative. The plots for males with "less than high school" education and females with "high school equivalent" education appear to have heteroskedastic trend lines negating a linear relationship. For both sex groups in the "less than high school" education category, there appears to be an increase in total activity around age 60. For females in the "high school equivalent" education category there appears to be a greater difference in total activity from younger ages to older ages than in the male "high school equivalent" category, indicating a greater spread in the data. The distribution of total activity in the "more than high school" category appears to be fairly similar for both sexes. 

```{r}

nhanes_hour <- nhanes %>%
  pivot_longer(cols = starts_with("min"), names_to = "minute", values_to = "activity") %>%
  mutate(
    minute = as.numeric(gsub("min", "", minute)),  
    hour = (minute - 1) %/% 60) %>%
  group_by(education, sex, hour) %>%
  summarize(mean_activity = mean(activity)) 

day_plot = 
  nhanes_hour %>% 
  ggplot(aes(x = hour, y = mean_activity, color = sex)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_grid(.~education) +
  labs(
    title = "24-Hour Activity Time by Education Level and Sex", x = "Hour", y = "Average Activity Time")

day_plot
```

Based on the three-panel plot showing the distribution of average activity levels by hour over a 24-hour period by education category and sex, we can see that average activity per hour follows a similar pattern for both sexes and education groups. In each education and sex category, average activity per hour increases in the late morning and afternoon, peaking around 10:00 am to 3 pm and decreasing from there. Average activity by hour appears to be higher for females in all education groups, however, the activity levels are most similar between sexes in the "less than high school" education category. Based on this plot we can conclude that individuals in all education categories follow similar activity patterns based on the time of day. It also tells us that individuals with higher levels of education have higher average activity levels throughout the day when compared to those with lower levels of education. 
